---
title: "TURNT - Water Nutrients exploration"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    smooth_scroll: true 
    fig_width: 10
    code_folding: hide  
date: '2022-12-05'
---
# Overview
This rmarkdown explores the water nutrient data associated with the Mo'orea 2021 Turbinaria nutrient addition (and remineralization) experiment: *TURNT*.

There were three treatments:

 C = control, collected at 'normal' reef site, no nutrient addition
 
 N = nutrient, collected at 'normal' reef site site, osmocote added during experiment 
 
 R = recharge, collected at long-term osmocote addition site, osmocote added during experiment 



Purpose of this document:

  1) Organize water nutrient data 
  
  2) Compare nutrient values and ratios between treatments

  3) compare nutrient values and ratios to natural habitat
  
  
  
```{r}
library(ggplot2)
library(data.table)
theme_set(theme_minimal())

turnt_clrs = c(
  control = "#73d9f0",
  nutrient = "#f7b301",
  recharge = "#eb773e",
  blank = "gray"
  )
```


```{r}
# read in
nut_file = "../data/nutrient_data/water/clean_nutrients.csv"
meta_file = "../data/sample_data/TURNT_sample_sheet - samples.csv"

alt_met_file = "../data/sample_data/TURNT_sample_sheet - water_nutrient.csv"



nut = fread(nut_file)

meta = fread(meta_file)
setnames(meta, "MooreaCode", "Sample")
meta = meta[!is.na(Sample)]

# time point comparisons during exudation
alt_meta = fread(alt_met_file)
setnames(alt_meta, "nutrient_id", "Sample")


# nutrient variables
nut_list = list("N",
            "P",
            "Phosphate",
            "Ammonia",
            "Silicate",
            "N+N"
            )


```

Check the exudation time point nutrients. Did fertilizing with osmocote produce the desired effect?


```{r}
trt_nut = merge(alt_meta, nut, all.x = T, all.y =F)
trt_nut = trt_nut[!is.na(P),]


trt_plot = function(y_var, dat = trt_nut) {
  
  ggplot(dat, aes_string(x = "time_point_hr",
                  y = y_var,
                  color = "treatment",
                  group = "treatment"))+
  geom_point(size = 3, pch = 21)+
  geom_line(alpha = 0.7, size = 2)+
  scale_color_manual(values = turnt_clrs)

}

# Nutrients

lapply(nut_list, trt_plot)


```


Check exudation data 

```{r}
exp_nut = merge(meta, nut, all.x =F, all.y = F)


exp_nut_plot = function(y_var, dat = exp_nut){
 
  
  means = tapply(dat[[y_var]],
                 dat$treatment,
                 mean)
  
  dat$mean_val = means[dat$treatment]
  
  ggplot(dat, aes_string(x = "treatment",
                         y = y_var,
                         color = "treatment",
                         group = "treatment"))+
    geom_jitter(size = 3, pch = 21)+
    stat_summary(fun = "mean", geom = "point", pch = 19, size = 3)+
    scale_color_manual(values = turnt_clrs)+
    labs(title = unique(dat$stage))
   
}

# exudation1
lapply(nut_list, exp_nut_plot, dat = exp_nut[stage == "exudation_1",])

# exudation 2
lapply(nut_list, exp_nut_plot,  dat = exp_nut[stage == "exudation_2",])

# remin T0
lapply(nut_list, exp_nut_plot,  dat = exp_nut[stage == "remin_T0",])

# remin TF
lapply(nut_list, exp_nut_plot,  dat = exp_nut[stage == "remin_TF",])


```



